HO unification from object language to meta language

### 2024-03-31 paper/main.tex
update PBs 6 and 7
reword 6.0 and 6.1

\section{Handling of \maybeeta}\label{sec:eta}
\marginpar{TODO: fix syntax}
$\eta$-reduction is an equivalence relation where a term of the form
$\lambda x.t \appsep x$ can be converted to $t$ any time $x$ does not occur as a
free variable in $t$. We call $\lambda x. t \appsep x$ the $\eta$-expansion of $t$. 
\printAlll
  {{{\lambda x. X \appsep x, f}}}
  {{{\lambda x. A_x, f}}}
  {{{X,A,1}}}
  {{}}
Following the implementation of \elpiIn{comp} given in
\cref{sec:compilation}, the unification problem $\mathbb{P}_1$ given 
above, is compiled into the unification problem $\mathbb{T}_1$. 
We can note that all the terms appearing in $\mathbb{P}_1$ are
unifiable by \Uo, but, despite this, \Ue, is not able to solve $\mathbb{T}_1$:
its left and right hand sides have both different rigid head:
$\lambda x.A'_x$ corresponds to the \Ho term \elpiIn{lam x\ app[con "f", x]} 
and $f$ corresponds to \elpiIn{con"f"}. 
This failure is motivated by the fact that \elpiIn{flam x\ fapp [fuva A, x]}
is a term belonging to \maybeeta.

In order to guarantee \cref{prop:simulation}, we need to modify the way terms are
compiled. The goal is to recognize
every \maybeeta subterm $t$ and replace it with fresh \Ho variables $v$. 
This connection between the variable $v$ and the subterm $t$ is stored in what we
call \linketa which is an object with the following type

@@

\noindent
where, as sketched in \cref{sec:lang-spec}, the term on the left hand side
(\lhs) is linked with its left counterpart (\rhs).

\linketa are added in the link store (\linkStore) and activated when special
conditions are satisfied on \lhs or \rhs. These link activations are managed by
extending the \elpiIn{progress1} predicate (see \cref{sec:execution}).
We claim that \linketa progression does not contradict \cref{inv:linklhs} and we
add the following invariant:

\begin{invariant}[\linketa \rhs]
  The \rhs of any \linketa in \linkStore has the shape $\lambda x.t_x$
  where $t_x$ is a \maybeeta term and $x$ is free in $t$.
  \label{inv:link-eta-right}
\end{invariant}

In the next three subsections we explain how we detect \maybeeta terms, how we
compile them and how the generated \linketa are activated during the \elpiIn{progress}.
Moreover, we provide justification for why \cref{inv:linklhs,inv:link-eta-right} remain true.

@@

\begin{lemma}[\maybeeta detection]\label{lem:maybeeta}
  A $\beta$-normal term
  $\lambda x_1 \ldots x_n.t \in \maybeeta$ (as per \cref{def:maybeeta})
  % where each $x_i$ occurs in $t$ is a
  %\maybeeta if
  %either: 
  if one of the following holds:
  \begin{enumerate}
    \item $t$ is a constant or variable applied to arguments
      $l_1 \ldots l_m$ such that 
      $m \geq n$ and for every $i$ such that $1 \leq i \leq m-n$
      the term  $l_i$
      \reduceto{} $x_i$, and
      no $x_i$ \occursrigid{} in $l_1 \ldots l_{m-n-1}$;
    \item $t$ is a
      unification variable with scope $W$ and
      for each $x_i$ there exists a $w_j \in W$ such that $w_j$
      \reduceto{} $x_i$.
  \end{enumerate}
\end{lemma}
\begin{proof}[Proof sketch]
Follows from \cref{def:occrigid} and \cref{lem:reduceto}
\end{proof}

\noindent
Remark that the converse of \cref{lem:maybeeta} does not hold: 
there exists a term $t$ satisfying the criteria (1) above
that is not in $\maybeeta$, i.e.
there exists no substitution $\rho$ such that $\rho t$ is an
$\eta$-expansion. A simple counter example is
$\lambda x. f \appsep (A \appsep x) \appsep (A \appsep x)$
since $x$ does not \occurrigid{} in the first argument
of $f$,
and the second argument of $f$ \reduceto{} $x$.
In other words $A \appsep x$
may either use or discard $x$, but our analysis does not
take into account that \emph{the same
term} cannot have two contrasting behaviors.

@@

Thanks to the \elpiIn{maybe-eta} predicate, we can detect ``$\eta$-problematic''
terms and, consequently replace them with fresh \Ho unification variables at
compilation time. The code below illustrate how this relation is used to for
term compilation.

@@

This rule, to be inserted just before rule~\ref{rule:complam} from the code in
\cref{sec:compilation}, verifies if the \Fo{} term $t$ received in entry is a
\maybeeta. Let $\lambda t'$ be the compiled version of $t$, then the fresh
variable \elpiIn{A} returned as the new \Ho term as in scope all the free names
in $t'$. The critical part of this compilation is the creation of the \linketa,
which links the variable \elpiIn{A} with $t$. This link creation enforce
\cref{inv:linklhs} and \cref{inv:link-eta-right}, since \lhs is a trivially a
variable and the \rhs is a term $t$ starting with the \elpiIn{lam} constructor where
$x$ is free in $t$ otherwise $t$ would not have been detected as a \maybeeta.

@@

Once a \linketa is activated, it can be removed from \linkStore,
otherwise, the link is kept for a further iteration of \elpiIn{progress}. Note
that this link progression enforce
\cref{prop:nf,inv:linklhs,inv:link-eta-right}: we never commit a term in
the \Ho substitution, since we make unification only when we know that the 
terms are no more \maybeeta, and when \lhs is no more a variable or \rhs is 
no more a \maybeeta, the link is removed from \linkStore.
% \printAll
%   {{\elpiIn{flam x\ fapp [fuva X, x] ~\Uo~fcon"f"}}}
%   {{A \Ue f}}
%   {{\mapping{X}{X'}{1}}}
%   {{\linketaM{}{A}{\lambda x.X'_{x}}}}
$$
\begin{array}{l}
  \mathbb{P} = \{~ \lambda x.X \appsep x \Uo f ~\}\\
  \mathbb{T} = \{~ A \Ue f ~\}\\
  \mathbb{M} = \{~ \mapping{X}{B}{1} ~\}\\
  \linkStore \;= \{~ \linketaM{}{A}{\lambda x.B_{x ~\}}}
\end{array}
$$

@@

In \cref{sec:basic-comp-limitations}, we have given two unification problems to
be run one after the other. In the following table we present the entry problem,
its compiled version, with the corresponding list of mapping and links.

% \printAll
%   {{\elpiIn{flam x\ flam y\ fapp [fuva X, y, x] == flam x\ flam y\ x},
%     \elpiIn{flam x\ fapp [fcon"f", fapp [fuva X, x], x] == fuva Y}}}
%   {{E0 == \lambda x.\lambda y.x,
%     E3 == E5}}
%   {{\mapping{Y}{E5}{0},
%     \mapping{X}{E2}{2}}}
%   {{\linketaM{c0}{E4_{c0}}{\lambda x.E2_{c0 x}},
%     \linketaM{}{E3}{\lambda x.(f~E4_{x}~x)},
%     \linketaM{}{E0}{\lambda x.E1_{x}},
%     \linketaM{c0}{E1_{c0}}{\lambda x.E2_{x c0}}}}
$$
{\arraycolsep=2pt \begin{array}{rrclrcl}
    \mathbb{P} = \{ &
       \lambda x.\lambda y.X \appsep y \appsep x & \Uo & \lambda x.\lambda y. x & \quad
       \lambda x. f \appsep (X \appsep x) \appsep x & \Uo&  Y ~\}\\
    \mathbb{T} = \{ &
       A &\Ue& \lambda x.\lambda y.x & \quad
       D &\Ue& F ~\}\\
    \mathbb{M} = \{ & \multicolumn{6}{l}{
      \mapping{Y}{F}{0} \quad
      \mapping{X}{C}{2} ~\}}\\
    \linkStore \;= \Big\{  &
      \multicolumn{6}{l}{ \begin{array}{ll}
          \linketaM{}{A}{\lambda x.B_{x}} & \quad
          \linketaM{x}{B_{x}}{\lambda y.C_{y x}}\\
          \linketaM{}{D}{\lambda x.(f~E_{x}~x)} & \quad
          \linketaM{y}{E_{y}}{\lambda w.C_{y w}}
      \end{array} ~\Big\}}
\end{array}}
$$

\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{E0,\lambda x.\lambda y.x},
    {E3,E5}}}
  {{{Y,E5,0},
    {X,E4,1},
    {X,E2,2}}}
  {{{,E3,\lambda x.(f\appsep E4_{x}\appsep x)},
    {,E0,\lambda x.E1_{x}},
    {x,E1_{x},\lambda y.E2_{y x}}}}

@@

The complete problem to run for resolution is now:
%
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{E0,\lambda x.\lambda y.x},
    {E3,E5}}}
  {{{Y,E5,0},
    {X,E2,2}}}
  {{{x,E4_{x},\lambda y.E2_{x y}},
    {,E3,\lambda x.(f\appsep E4_{x}\appsep x)},
    {,E0,\lambda x.E1_{x}},
    {x,E1_{x},\lambda y.E2_{y x}}}}

>>>>

\section{Handling of \maybeeta}\label{sec:eta}
$\eta$-reduction is an equivalence relation where a term of the form
$\lambda x.t \appsep x$ can be converted to $t$ any time $x$ does not occur as a
free variable in $t$. We call $t$ the $\eta$-contraction of
$\lambda x. t \appsep x$.

Following the compilation scheme of \cref{sec:compilation} the
unification problem $\mathbb{P}$ is compiled as follows:
%
\printAlll
  {{{\lambda x. X \appsep x, f}}}
  {{{\lambda x. A_x, f}}}
  {{{X,A,1}}}
  {{}}

\noindent
While the original problem should be solved by \Uo{} (by assigning $f$ to $X$)
\Ue{} is not able to solve the problem in $\mathbb{T}$:
its left and right hand sides have different rigid heads:
$\lambda x.A_x$ corresponds to the \Ho{} term
\elpiIn{lam x\ uva A [x]} 
while $f$ corresponds to
\elpiIn{con"f"}.

In order to guarantee \cref{prop:simulation} we detect
lambdas that can disappear by eta contraction (the \maybeeta set
introduced in \cref{sec:nutshell}); then 
we modify the way terms are compiled and move terms in \maybeeta
from  $\mathbb{T}$ to the \linkStore and put fresh unification variables
in their place.

@@

\noindent
As per \cref{inv:linklhs} the term on the left hand side
(\lhs) is a variable, and its right counterpart (\rhs) is the
term in \maybeeta and it has the following property:

\begin{invariant}[\linketa \rhs]
  The \rhs of any \linketa in \linkStore has the shape $\lambda x.t$
  and $t$ is not a lambda. 
  %where $t_x$ is a \maybeeta term and $x$ is free in $t$.
  \label{inv:link-eta-right}
\end{invariant}

\linketa are kept in the link store (\linkStore) during execution
and activated when some conditions hold on \lhs or \rhs.
Link activation is implemented by extending the \elpiIn{progress1}
predicate (defined in \cref{sec:execution}).

@@

\newcommand{\testmaybeeta}{\emph{maybe-eta}}
\begin{definition}[\testmaybeeta]\label{def:testmaybeeta}
  Given a $\beta$-normal term
  $s = \lambda x_1 \ldots x_n.t$, \testmaybeeta{} $s$ holds if any
  of the following holds:
  \begin{enumerate}
    \item $t$ is a constant or variable applied to arguments
      $l_1 \ldots l_m$ such that 
      $m \geq n$ and for every $i$ such that $1 \leq i \leq m-n$
      the term  $l_i$
      \reduceto{} $x_i$, and
      no $x_i$ \occursrigid{} in $l_1 \ldots l_{m-n-1}$;
    \item $t$ is a
      unification variable with scope $W$ and
      for each $x_i$ there exists a $w_j \in W$ such that $w_j$
      \reduceto{} $x_i$.
  \end{enumerate}
\end{definition}
\begin{lemma}[\maybeeta detection]\label{lem:maybeeta}
  If $t$ is a $\beta$-normal term and \testmaybeeta{} $t$ holds,
  then $t \in \maybeeta$.
\end{lemma}
\begin{proof}[Proof sketch]
Follows from \cref{def:occrigid} and \cref{lem:reduceto}
\end{proof}

\noindent
Remark that the converse of \cref{lem:maybeeta} does not hold: 
there exists a term $t$ satisfying the criteria (1) of
\cref{def:testmaybeeta} that is not in $\maybeeta$, i.e.
there exists no substitution $\rho$ such that $\rho t$ is an
$\eta$-expansion. A simple counter example is
$\lambda x. f \appsep (A \appsep x) \appsep (A \appsep x)$
since $x$ does not \occurrigid{} in the first argument
of $f$,
and the second argument of $f$ \reduceto{} $x$.
In other words $A \appsep x$
may either use or discard $x$, but our analysis does not
take into account that \emph{the same
term} cannot have two contrasting behaviors.

@@

% Thanks to the \elpiIn{maybe-eta} predicate, we can detect ``$\eta$-problematic''
% terms and, consequently replace them with fresh \Ho unification variables at
% compilation time. The code below illustrate how this relation is used to for
% term compilation.
The following rule is inserted just before rule~\ref{rule:complam} from the code in
\cref{sec:compilation}.

@@

\noindent
The rule verifies if the \Fo{} term $\lambda t$ in input is in
\maybeeta. Let $\lambda t'$ be the compiled version of $\lambda t$, then the fresh
variable \elpiIn{A} returned as the new \Ho term as in scope all the free names
in $t'$. The critical part of this compilation is the creation of the \linketa,
which links the variable \elpiIn{A} with $t$. This link creation enforce
\cref{inv:linklhs} and \cref{inv:link-eta-right}, since \lhs is a trivially a
variable and the \rhs is a term $t$ starting with the \elpiIn{lam} constructor where
$x$ is free in $t$ otherwise $t$ would not have been detected as a \maybeeta.

@@

Once a \linketa is activated, it can be removed from \linkStore,
otherwise, the link is kept for a further iteration of \elpiIn{progress}. Note
that this link progression enforce
\cref{prop:nf,inv:linklhs,inv:link-eta-right}: we never commit a term in
the \Ho substitution, since we make unification only when we know that the 
terms are no more \maybeeta, and when \lhs is no more a variable or \rhs is 
no more a \maybeeta, the link is removed from \linkStore.
% \printAll
%   {{\elpiIn{flam x\ fapp [fuva X, x] ~\Uo~fcon"f"}}}
%   {{A \Ue f}}
%   {{\mapping{X}{X'}{1}}}
%   {{\linketaM{}{A}{\lambda x.X'_{x}}}}
% $$
% \begin{array}{l}
%   \mathbb{P} = \{~ \lambda x.X \appsep x \Uo f ~\}\\
%   \mathbb{T} = \{~ A \Ue f ~\}\\
%   \mathbb{M} = \{~ \mapping{X}{B}{1} ~\}\\
%   \linkStore \;= \{~ \linketaM{}{A}{\lambda x.B_{x ~\}}}
% \end{array}
% $$

@@

In \cref{sec:basic-comp-limitations}, we have given two unification problems to
be run one after the other. In the following table we present the entry problem,
its compiled version, with the corresponding list of mapping and links.
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{Y,F,0},
    {X,E,1},
    {X,C,2}}}
  {{{,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {,A,\lambda x.B_{x}},
    {x,B_{x},\lambda y.C_{y x}}}}

@@

The complete problem to run for resolution is now:
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{Y,F,0},
    {X,C,2}}}
  {{{x,E_{x},\lambda y.C_{x y}},
    {,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {,A,\lambda x.B_{x}},
    {x,B_{x},\lambda y.C_{y x}}}}
